Computable Compressed Matrices\\Suplementary Material

### 2014-02-07 manuscript.tex
fiz a revisão do texto removendo todos os "\todo"s e acrescentando outros

In this paper, we present two methods for the lossless compression of 
(numerical) arrays. The methods involve the encoding of the numbers as strings 
of bits of variable length. The methods resemble the arithmetic 
coding\cite{bodden2007arithmetic} algorithm, but is cheaper to compute. We 
describe the process of compression and decompression, and study their 
efficiency under different applications. We also discuss the efficiency of the 
compression as a function of the distribution of the elements of the matrix. 
\todo{Inseri isso} Also presented are some results in relation to the 
traditional method of matrix allocation. We also consider the 
arithmetic operations to measure the efficiency of the proposed methodology.

@@

To maintain mathematical equivalence with the original data for any arithmetic 
operations, we need to maintain the structure of the matrix, ie, the ability 
to access any element given its row $i$ and column $j$ and also the numeric 
nature of its elements. In order to achieve compression, we decided to exploit 
inefficiencies in the conventional way matrices are allocated in memory. 
\todo{adicionei aqui} The analyzes presented in this article, in the first 
moment, are related to arrays of positive integers, but can be applied to 
integers and real numbers with some adaptations, which are presented in due 
course. \todo{removi isso The examples in this paper will be restricted to 
matrices with positive integer elements, but}

The compression method is as follows. Let $M_{r \times c}$ be a matrix, in which 
$r$ is the number of rows and $c$ the number of columns. Each element of this 
matrix, called $m_{ij}$, is a positive integer. In digital computers, all 
information is stored as binary code (base 2 numbers). However, the conventional 
way to store arrays of integers is on a memory block sequence of fixed size 
(power of 2 numbers of bit), one for each element. The maximum size of a block 
is equal to the word size of the processor, which for most current CPUs is 64 
bits. Some special number such as complex number may be encoded as two blocks 
instead of one. The size of the chunk of memory allocated to each number will 
determine their maximum  size (for integers) or their precision (for 
floating-point numbers). So for matrix $M$, the total memory allocated, 
assuming chunks of 64 bits, is given by $\mathcal{B} = r \times c \times 64$.

@@

\begin{equation} \label{eq:01}
 b(\sup M) \approx \begin{cases}
	1, &  \text{ if } \sup M \in \{0,1\} \\ 
	\lfloor \log_2(\sup M)  \rfloor + 1,  & \text{ if } \sup M > 1  
	\end{cases}
\end{equation}\todo{Será que não devemos trocar $\approx$ por = aqui? Note que 
o número de bits não é aproximado e sim o exato.}

@@

 We used a total of 87 bits to store matrix $m$ with the VLB method instead of 
80 bits using the SM method. However, as shall be seen later, the VLB method 
will be the most efficient for most matrices. \todo{adicionei} With the method 
VLB the resulting matrix is smaller than the original matrix.

@@

As we see in \ref{eq:15}, $\eta_1$ does not depend on size of the matrix, only 
on the bit-length of $\max M$. If $b(\max M)=64$, $\eta_1$ is 0, ie, no 
compression is possible. On the other extreme, if the matrix is composed 
exclusively of 0s and 1s, maximal compression is achievable, 
$\eta_1=0.9843$.\todo{Alterei a proporção para 0.98, pois é o maior valor que 
pode ser obtido.}

@@

For this method, the highest value obtained for the data compression is 0.9687, 
when b = 1 and k = 1. However, the lower value is defined by a range of values, 
​​such that the value of $k = \frac{\sum_{i=1}^{g}  b_i \times f_i}{rc} - 64$. 
\todo{Adicionei esse trecho para ficar igual ao SM}

@@

As we have shown, the VLB method is more effective compressing most integer 
datasets up to 64 bits in size. This is due to its ability to exploit the 
variance in the data set and reduce the waste of bits in the representation of 
some numbers. In specific cases, where the variance in the data null or too 
small, method I will be more efficient. As a matter of fact, for matrices where 
all elements have the same bit-length, SM method  will always be better, 
regardless of bit-length (Figures \ref{fig20} and \ref{fig21}). There are only 
exception, if for bit-length 64, where neither method is able to compress the 
data. \todo{Alterei a última frase}

@@

To determine the best compression method to apply, it's necessary to inspect the 
distribution of bit-lengths of matrix elements. When matrix elements are small 
or have nearly-constant bit-length, the SM Method is better\todo{alterei de 
best para better}, otherwise, the VLB method should be chosen.

@@

The efficiency of the VLB method is influenced by the relative size of the 
bit-length groups. In this first example, we considered only two groups, each 
comprised of half the matrix elements. Let's now vary the relative frequency of 
the groups, $\frac{f_i}{rc}$, while sticking to two groups. Let's 
also \todo{Acho que tinha que tirar esse Let's also, talvez colocar Now, we are 
going to} assume that $\frac{f_i}{rc}$ is a good approximation to the 
probability of a given bit-length in a matrix, which we will denote by $p_i$.

@@

\paragraph{Numerical Estimates:}
The results for this simulation can be seen in \ref{tab:06}. Note that 
increasing the number of bits to represent numbers increases, there is a loss of 
efficiency in the compression process. In this case, we did not simulate for 
$\lambda=64$ since a large portion of the samples would fall above the maximum 
bit-length, but \todo{coloquei ", but", aqui} we are considering for this 
analysis.

@@

\section*{Library Implementation }

The methodology employed in this this paper includes the SM method, which was implemented in a library for Fortran. The 
overloaded operations were assignment, addition, subtraction, multiplication, transpose, and maximum of a matrix. 
Although only few significant operations of matrices presented ovearload, any other operation can be performed, since 
the methods for inserting and collecting an element in a matrix of bitstring were implemented. With these two methods, 
as long as they are properly adapted into algorithms, it becomes possible to implement any other desired operation.

The method was also extended to allow for the allocation of real and integer numbers, besides positive integers, inside 
the matrix bistring. Some adjustments were made so that the other groups of numbers could be represented. In order to 
represent the integers with signs, it is necessary to use two's complement, as it is applied to thein accordance to 
traditional methods.

The SM method uses a range of bits to store the values in such a way that
the size or number of bits used is important to represent the highest value in the bitstring matrix. The adaptation 
consists in determining the largest value to be stored considering the absolute value of the number. That done,
when the number is negative, it is necessary to rewrite it since, due to the sign, this will be written in the form of 
two's complement. The purpose of this conversion is to optimize library processing. Take the following as an example: 
the number 5, when represented in a 32-bit integer, is defined as:

@@

\section*{Library Application}

>>>>

In this paper, we present two methods for the lossless compression of 
(numerical) arrays. The methods involve the encoding of the numbers as strings 
of bits of variable length. The methods resemble the arithmetic 
coding\cite{bodden2007arithmetic} algorithm, but is cheaper to compute. We 
describe the process of compression and decompression, and study their 
efficiency under different applications. We also discuss the efficiency of the 
compression as a function of the distribution of the elements of the 
matrix, some results comparing our method with the traditional method of matrix 
allocation, and arithmetic operations as a benchmark measure the efficiency of 
operating with bitstring-compressed matrices.

@@

To maintain mathematical equivalence with the original data for any arithmetic 
operations, we need to maintain the structure of the matrix, ie, the ability 
to access any element given its row $i$ and column $j$ and also the numeric 
nature of its elements. In order to achieve compression, we decided to exploit 
inefficiencies in the conventional way matrices are allocated in memory. 
 The analyses presented in this article concern arrays of positive integers, 
but can be applied to signed integers and real numbers with few adaptations.

The compression method is as follows. Let $M_{r \times c}$ be a matrix, in which 
$r$ is the number of rows and $c$ the number of columns. Each element of this 
matrix, called $m_{ij}$, is a positive integer. In digital computers, all 
information is stored as binary code (base 2 numbers). However, the conventional 
way to store arrays of integers is on a memory block sequence of fixed size 
(power of 2 number of bits), one for each element. The maximum size of a block 
is equal to the word size of the processor, which for most current CPUs is 64 
bits. Some special number such as complex number may be encoded as two blocks 
instead of one. The size of the chunk of memory allocated to each number will 
determine their maximum  size (for integers) or their precision (for 
floating-point numbers). So for matrix $M$, the total memory allocated, 
assuming chunks of 64 bits, is given by $\mathcal{B} = r \times c \times 64$.

@@

\begin{equation} \label{eq:01}
 b(\sup M) = \begin{cases}
	1, &  \text{ if } \sup M \in \{0,1\} \\ 
	\lfloor \log_2(\sup M)  \rfloor + 1,  & \text{ if } \sup M > 1  
	\end{cases}
\end{equation}\todo{Será que não devemos trocar $\approx$ por = aqui? Note que 
o número de bits não é aproximado e sim o exato.}

@@

 We used a total of 87 bits to store matrix $m$ with the VLB method instead of 
80 bits using the SM method. However, as shall be seen later, the VLB method 
will be the most efficient for most matrices. With the VLB 
method the resulting matrix is smaller than the original matrix.

@@

As we see in \ref{eq:15}, $\eta_1$ does not depend on size of the matrix, only 
on the bit-length of $\max M$. If $b(\max M)=64$, $\eta_1$ is 0, ie, no 
compression is possible. On the other extreme, if the matrix is composed 
exclusively of 0s and 1s, maximal compression is achievable, 
$\eta_1=0.9843$.

@@

For this method, the highest value obtained for the data compression is 0.9687, 
when b = 1 and k = 1. However, the lower value is defined by a range of values, 
​​such that the value of $k = \frac{\sum_{i=1}^{g}  b_i \times f_i}{rc} - 64$.

@@

As we have shown, the VLB method is more effective compressing most integer 
datasets up to 64 bits in size. This is due to its ability to exploit the 
variance in the data set and reduce the waste of bits in the representation of 
some numbers. In specific cases, where the variance in the data null or too 
small, method I will be more efficient. As a matter of fact, for matrices where 
all elements have the same bit-length, SM method  will always be better, 
regardless of bit-length (Figures \ref{fig20} and \ref{fig21}). There are only 
one exception: for bit-length 64, where neither method is able to compress 
the data.

@@

To determine the best compression method to apply, it's necessary to inspect the 
distribution of bit-lengths of matrix elements. When matrix elements are small 
or have nearly-constant bit-length, the SM method is better, otherwise, the VLB 
method should be chosen.

@@

The efficiency of the VLB method is influenced by the relative size of the 
bit-length groups. In this first example, we considered only two groups, each 
comprised of half the matrix elements. Let's now vary the relative frequency of 
the groups, $\frac{f_i}{rc}$, while sticking to two groups. Here, we assume 
that $\frac{f_i}{rc}$ is a good approximation to the 
probability of a given bit-length in a matrix, which we will denote by $p_i$.

@@

\paragraph{Numerical Estimates:}
The results for this simulation can be seen in table \ref{tab:06}. Note that 
increasing the number of bits to represent numbers increases, there is a loss of 
efficiency in the compression process. In this case, we did not simulate for 
$\lambda=64$ since a large portion of the samples would fall above the maximum 
bit-length.

@@

\section*{Implementation }\todo{acho que esta seção toda deve ir para o 
material suplementar.}

This paper proposes the bitstring-compression methodologies. The SM 
method was implemented as a Fortran library which is available under an 
open-source license\todo{colocar o link para o repositório aqui}. The 
implementation extends the standard Matrix type of Fortran, overloading 
operators such as assignment, addition, subtraction, multiplication, transpose, 
and maximum of a matrix. 
Although only few significant matrix operations have been ovearloaded in this 
example implementation, any other operation can be performed, since 
the methods for inserting and collecting an element in a matrix of bitstring 
were also implemented. With these two methods, as long as they are properly 
adapted into algorithms, it is possible to implement any other desired 
operation.

The SM method was also extended to allow for the allocation of real and 
signed integer matrices, besides positive integer ones. Some adjustments were 
made so that the other types of numbers could be represented. In order to 
represent the signed integers, it was necessary to use two's complement, as it 
is applied to thein accordance to traditional methods\todo{acho que esta 
explicação ficou muito solta. Não precisamos explica a implementação no 
artigo, podemos colocar na documentação do código}.

The SM method uses a fixed number of bits to store the values in such a way that
the size or number of bits used is important to represent the highest value in the bitstring matrix. The adaptation 
consists in determining the largest value to be stored considering the absolute value of the number. That done,
when the number is negative, it is necessary to rewrite it since, due to the sign, this will be written in the form of 
two's complement. The purpose of this conversion is to optimize library processing. Take the following as an example: 
the number 5, when represented in a 32-bit integer, is defined as:

@@

\section*{Library Application}\todo{Também deve ir para o material suplementar}
